### 2015-05-13 whoop.tex
draft sections for watchdog-based techniques

@@

\medskip\noindent\textbf{Identifying Set of Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.

@@

\medskip\noindent\textbf{Instrumentation and Analysis }
%
Next, \whoop instruments the program with global variables for lockset analysis and race-checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:

@@

\whoop proceeds by instrumenting each entry point call-graph for lockset analysis by replacing each call to a locking or unlocking function with a \whoop-specific lock and unlock functions. This means that Example~\ref{fig:smack_translation} will become:

\begin{boogie}
procedure ep1(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    $p0 := dev;
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, true);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    $M.0[$p4] := 4;
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, false);
    return;
}
\end{boogie}

The \texttt{\_UPDATE\_CLS\_\$ep1()} is a special function that updates the CLS global variable to either true or false, if the lock is held or released by \texttt{ep1} accordingly:

\begin{boogie}
procedure {:inline 1} _UPDATE_CLS_$ep1(lock: int,
    isLocked: bool);
  modifies lock$0_in_CLS_$ep1;

implementation {:inline 1} _UPDATE_CLS_$ep1(
    lock: int, isLocked: bool) {
  _UPDATE:
    lock$0_in_CLS_$ep1 := (if lock == lock$0 then
      isLocked else lock$0_in_CLS_$ep1);
  return;
}
\end{boogie}

Next, \whoop replaces every write and read access in each entry point call-graph with a \whoop-specific function:

\begin{boogie}
procedure ep1(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    $p0 := dev;
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, true);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    call _WRITE_LS_$M.0_$ep1($p4);
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, false);
    return;
}
\end{boogie}

The \texttt{\_WRITE\_LS\_\$M.0\_\$ep1()} is a special function that computes the intersection of the CLS and the lockset of \texttt{\$M.0} (see Section~\ref{bg:lockset}). It also updates the write-set of \texttt{\$M.0} to true, because there was a write access by \texttt{ep1}:

\begin{boogie}
procedure {:inline 1} _WRITE_LS_$M.0_$ep1(
    ptr: int);
  modifies lock$0_in_LS_$M.0_$ep1,
    WRITTEN_$M.0_$ep1;

implementation {:inline 1} _WRITE_LS_$M.0_$ep1(
    ptr: int) {
  _WRITE:
    goto anon1_Then, anon1_Else;

  anon1_Then:
    assume {:partition} WATCHED_ACCESS_$M.0 ==
      ptr && DEVICE_IS_REGISTERED_$ep;
    lock$0_in_LS_$M.0_$ep := lock$0_in_CLS_$ep && lock$0_in_LS_$M.0_$ep;
    WRITTEN_$M.0_$ep := true;
    return;
  anon1_Else:
    assume {:partition} !(WATCHED_ACCESS_$M.0 ==
      ptr && DEVICE_IS_REGISTERED_$ep);
    return;
}
\end{boogie}

The \texttt{\_READ\_LS\_\$M.0\_\$ep1()} updates the lockset of \texttt{\$M.0} and the read-set accordingly for a read access.

\whoop continues by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialization}, which combines the two arbitrary threads in a single sequential pair. This is achieved by creating a checker function that calls the two instrumented entry points and then performs race-checking assertions. This process repeats until all possible pairs of entry points have been sequentialized. To be sound, each time an entry point performs a read access to a shared memory location, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}). This over-approximates any effects from all the unmodeled threads on the shared state of the pair. Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.

After each checker function calls a pair of entry points, it asserts that each shared memory location is protected by at least one common lock (see Section~\ref{bg:lockset}). The assertion check is triggered iff the memory location was accessed by both entry points and at least one of the accesses was a write. We do not check for read-read races, which are inherently benign. The following is an example of a checker function for entry points \texttt{ep1} and \texttt{ep2}:

\begin{boogie}
procedure check$ep1$ep2(dev: int);
  requires !lock$0_in_CLS_$ep1;
  requires !lock$0_in_CLS_$ep2;
  requires lock$0_in_LS_$M.0_$ep1;
  requires lock$0_in_LS_$M.0_$ep2;
  requires !WRITTEN_$M.0_$ep1;
  requires !WRITTEN_$M.0_$ep2;
  requires !READ_$M.0_$ep1;
  requires !READ_$M.0_$ep2;
  modifies lock$0_in_CLS_$ep1, lock$0_in_CLS_$ep2,
    lock$0_in_LS_$M.0_$ep1, lock$0_in_LS_$M.0_$ep2,
    WRITTEN_ $M.0_$ep1,WRITTEN_$M.0_$ep2,
    READ_$M.0_$ep1, READ_$M.0_$ep2;

implementation check$ep1$ep2(dev: int) {
  _CHECK:
    call ep1(dev);
    call ep2(dev);
    assert (WRITTEN_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
      || (WRITTEN_$M.0_$ep1 && READ_$M.0_$ep2)
      || (READ_$M.0_$ep1 && WRITTEN_$M.0_$ep2)
      ==> lock$0_in_LS_$M.0_$ep1 &&
      lock$0_in_LS_$M.0_$ep2;
    return;
}
\end{boogie}

>>>>

\PDComment{Mention this somewhere (it was in my previous implementation text)? Two-thread reduction is not a new idea, it has been used before in GPU kernel verification~\cite{bardsley2014engineering} and in model checking of cache coherence protocols~\cite{mcmillan1999verification}.}

@@

\medskip\noindent\textbf{Identifying Set of All Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.

@@

\medskip\noindent\textbf{Watchdog Race-Checking }
%
Next, \whoop instruments the program with global variables for lockset analysis and race-checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:

@@
